---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bper: Bayesian Prediction for Ethnicity and Race

<!-- badges: start -->

<!-- badges: end -->

This package provides functions for predicting an individual's race/ethnicity given their first name, last name, geo-location, political party, age, gender, and address characteristics. The method is based on a Naive Bayes classification algorithm which incorporates known ethnorace distributions over the observed characteristics to generate posterior probabilities for each individual.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bwilden/bper")
```

## Usage

#### Step 1

Start by loading the `bper` package and preparing your input dataframe containing individuals whose ethnorace you wish to predict. The following columns are supported:

-   `first_name`. All names should be uppercase.
-   `last_name`. Remove all punctuation (e.g. O'NEAL becomes ONEAL) and all spaces to increase chances of finding a match. All names should be uppercase.
-   `birth_year`. Currently only individuals born between the years 1911 and 2010 will match.
-   `female`. A binary 1/0 indicator for female or not-female.
-   `party`. Options are "DEM" for Democrat, "REP" for Republican, and "UNA" for Independents or other political parties.
-   `apartment`. A binary 1/0 indicator for whether the individual lives in multi-unit housing or not.
-   `zip`. 5 digit ZIP Code.
-   `block`. 15 digit complete US Census block code. This is comprised of a 2-digit State code + 3-digit County code + 6-digit Census Tract code + 4-digit Census Block code.

If any of the above columns are not present in your input dataframe, a column of NA's will be appended so that the `predict_race` function works. See example below. The more input information the into the algorithm, the better the predictions!

```{r example}
library(bper)

example_persons
```

#### Step 2

The `predict_race` function works by merging nationwide ethnorace distribution data into the original data frame to perform the computations. This data is stored externally and must be downloaded and then loaded into your global environment before running `predict_race`. Do this with:

```{r}
load_bperdata(download = TRUE, save_files = FALSE)
```

For more details see `?load_bperdata` and <https://github.com/bwilden/bperdata.> Note: you do not need to install the `bperdata` package.

#### Step 3

Now you are ready to run `predict_race`!

```{r}
predict_race(example_persons)
```
